Biological Sequence Analysis {#sec:sequence-analysis}
============================
Strings in R {#sec:strings-in-r}
------------
The data type known as "strings" in most programming languages is
generally dealt with using character vectors in R, as we have seen so
far. Strings are fundamental data structures in pretty much any
context in computer science, but they are of special importance in
computational biology since they provide a natural representation for
biological sequences, be they DNA, RNA, or protein.

A few of the basic operations you can perform with character vectors
in R include:

### Querying string length {#sec:querying-string-length}
Unlike most other languages, the standard R function for finding out
how long a string---or, more generally, vector of strings, since R
sees a single string as just a character vector of length 1---is not
named length or any variant thereof, but instead `nchar`:

```{r}
# this string is a vector of length one
length("This is a string of 58 characters but a vector of length 1")
```

```{r}
nchar("This is a string of 58 characters but a vector of length 1")
```

```{r}
stringVec = c("string 1", "string 2 is longer")
nchar(stringVec)  ## returns integer vector
```

### Concatenating strings {#sec:concatenating-strings}
Using `paste` or `paste0`:

```{r}
s = paste("long", "horn")
s
```

```{r}
s = paste0("long", "horn")
s
```

`paste0` is really just `paste` with the `sep`arator set to the
empty string instead of a single space:

```{r}
paste("long", "horn", sep="")   ## same as paste0
```

```{r}
paste("long", "horn", sep="_SEPARATOR_")
```

You can `paste` together as many strings as you want:

```{r}
paste("abc", "def", "gh", "iklm", sep="_:_")
```

`paste` and `paste0` can both be used with character `vector`s
of length > 1 as well; the best way to understand what they do in
such cases is to study some examples:

```{r}
 ## can use the collapse argument to paste a single vector together:
# this is useful if we were trying to paste together a bunch of string vectors 
s = paste(c("long", "horn"), collapse="")
s
```

```{r}
 ## compare to default (no collapse argument provided) behavior:
paste(c("long", "horn"))
```

```{r}
 ## or
paste(c("long", "horn"), "suffix")
```

```{r}
 ## or even:
paste(c("long", "horn"),
      c("suffix1", "suffix2"))
```

### Taking substrings {#sec:taking-substrings}
Using `substr`:

```{r}
s = "longhorn"
substr(s, 1, 4)         ## returns first 4 characters of s
```

```{r}
substr(s, 2, 3)         ## returns substring with only 2 characters
```

```{r}
substr(s, 5, nchar(s))  ## returns suffix of s starting at 5th character
```

```{r}
substr(stringVec, 3, nchar(stringVec))
```

### String splitting {#sec:string-splitting}
String splitting in R can be done using `strsplit`. Because strings
are intrinsically embedded in character vectors in R, the output of
`strsplit` is provided as a nested `list` of `vectors`, with one
element of the outer `list` for each element of the input character
`vector`:

```{r}
stringVec
strsplit(stringVec, split="ng")
length(strsplit(stringVec, split="ng"))
```

```{r}
strsplit(stringVec, split="ng")[[2]]
```

```{r}
# strsplit(s)
strsplit(s, split="ng") ## still a list of vectors!
```

```{r}
#  strsplit returns a list  of same length as the input vector
strsplit(s, split="ng")[[1]]
```

```{r}
strsplit(s, split="ng")[[1]][[2]]
```

### Pattern matching {#sec:pattern-matching}
Using `grepl`: logical

```{r}
grepl("string", stringVec)
```

```{r}
grepl("longer", stringVec)
```

```{r}
grepl(s, stringVec)     ## "longhorn" not in either element of stringVec
```

### Pattern replacement {#sec:pattern-replacement}
Using `gsub` (for global substitute):

```{r}
gsub("string", "longhorn", stringVec)
```

```{r}
gsub("long", "stringi", stringVec)
```
